---
title: "Getting started with anovapowersim"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with anovapowersim}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)
```

`anovapowersim` primarily simulates power for balanced factorial ANOVA designs.
Specify the factors/levels, the term of interest, and a target partial eta
squared. `anovapowersim` generates default term-specific cell means, simulates
datasets, refits the ANOVA, and estimates power. An experimental interface for
one fixed unbalanced design is described below.

```{r setup, message=FALSE}
library(anovapowersim)
```

```{r load-precomputed-results, include=FALSE}
vignette_results_path <- system.file(
  "extdata",
  "anovapowersim-vignette-results.rds",
  package = "anovapowersim"
)
if (!nzchar(vignette_results_path)) {
  vignette_results_path <- file.path(
    "..",
    "inst",
    "extdata",
    "anovapowersim-vignette-results.rds"
  )
}
vignette_results <- readRDS(vignette_results_path)
```

## Search for the required sample size

The easiest way to get your required sample size is to use `power_n()` to
search for the sample size needed to reach the requested `power`.

This example is a 2 x 2 mixed design with one between-subjects factor
(`cond`) and one within-subject factor (`stim`).

We specify that we are interested in the `cond:stim` interaction, and that we want to have 90% power to detect a partial eta squared of 0.14. `power_n()` will search for the required sample size per between-subject cell, so `n = 17` gives total `N = 34`.

```{r adaptive-code, eval=FALSE}
power_n(
  between = c(cond = 2), # cond has 2 levels
  within = c(stim = 4), # stim has 4 levels
  term = "cond:stim",
  target_pes = 0.14,
  alpha = 0.05,
  power = 0.90,
  n_sims = 1000, # use 5000+ for a more precise estimate
  seed = 123 # for reproducibility
)
```

```{r adaptive-output, echo=FALSE}
vignette_results$adaptive
```

Note: here we use 1000 simulations for a quick example, but the package defaults to 10000 simulations for more precise estimates.

The output table uses compact column names: `n_per_cell` is the sample size per
between-subject cell, `total_n` is the full sample size, `num_df` and `den_df`
are the ANOVA degrees of freedom, `ncp` is the noncentrality parameter,
`power_calc` is the noncentral F power calculation, and `power_sim` is the
simulation estimate.

## Fixed-sample achieved power and sensitivity

`power_achieved()` and `power_sensitivity()` are experimental and available
only in the development version of `anovapowersim`. Their APIs and reporting
formats may change.

When sample size is already fixed, `power_achieved()` estimates simulated
power for a specified partial eta squared. The `n` argument is subjects per
between-subject cell, or total N for a purely within-subject design.

```{r achieved-power, eval=FALSE}
power_achieved(
  between = c(cond = 2),
  within = c(stim = 2),
  term = "cond:stim",
  target_pes = 0.14,
  n = 20,
  n_sims = 1000,
  seed = 123
)
```

Use `power_sensitivity()` when the effect size is unknown. It searches for the
smallest explicitly simulated partial eta squared that reaches the requested
power at the fixed sample size.

```{r sensitivity, eval=FALSE}
power_sensitivity(
  between = c(cond = 2),
  within = c(stim = 2),
  term = "cond:stim",
  n = 20,
  power = 0.90,
  n_sims = 1000,
  pes_tol = 0.001,
  seed = 123
)
```

When you want to skip simulations, use `power_achieved_calc()` and
`power_sensitivity_calc()`. These experimental functions use calculated
noncentral-F power and accept a planned nonsphericity correction through
`epsilon`.

```{r fixed-sample-calculations, eval=FALSE}
power_achieved_calc(
  between = c(cond = 2),
  within = c(stim = 3),
  term = "cond:stim",
  target_pes = 0.14,
  n = 20,
  gpower = TRUE,
  epsilon = 0.80
)

power_sensitivity_calc(
  between = c(cond = 2),
  within = c(stim = 3),
  term = "cond:stim",
  n = 20,
  power = 0.90,
  gpower = TRUE,
  epsilon = 0.80
)
```

## Power for unbalanced designs

`power_unbalanced()` is an experimental, development-version-only function for
the less common case where the exact allocation, population means, and
standard deviations are already known. It simulates that one fixed design; it
does not search over sample sizes and does not return calculated power.

Define every cell in the complete factorial design with `cell_design()`. The
factor values identify the cell, `n` is the number of subjects in its
between-subject group, `m` is its population mean, and `sd` is its marginal
population standard deviation. For repeated-measures designs, the same `n`
must therefore appear on every within-subject row belonging to a given
between-subject group.

```{r unbalanced-design, eval=FALSE}
unbalanced_design <- cell_design(
  group = "control",   time = "pre",  n = 22, m = 10.0, sd = 2.0,
  group = "control",   time = "post", n = 22, m = 11.0, sd = 2.2,
  group = "treatment", time = "pre",  n = 31, m = 10.1, sd = 2.4,
  group = "treatment", time = "post", n = 31, m = 12.4, sd = 2.8
)
```

Here, the means define the complete effect pattern, including any main effects
and interactions. The standard deviations can differ across groups and
within-subject cells. Name the repeated-measures factors in `within`; all other
factors are treated as between-subject factors.

For a within-subject design, use `unbalanced_covariance()` to define the
correlations among repeated measurements. It deliberately has no standard
deviation argument: `power_unbalanced()` combines the correlations with the
cell-specific `sd` values above to construct the covariance matrix separately
for each between-subject group. With one within-subject factor, correlation
pair names use its level names, such as `"pre:post"`. With multiple within-
subject factors, join the level combinations with underscores, as described in
the nonsphericity section below.

```{r unbalanced-power, eval=FALSE}
power_unbalanced(
  design = unbalanced_design,
  within = "time",
  term = "group:time",
  covariance = unbalanced_covariance(
    default_correlation = 0.5,
    correlations = c("pre:post" = 0.7)
  ),
  n_sims = 5000,
  parallel = TRUE,
  seed = 123
)
```

The result reports simulated power, the partial eta squared from a
deterministic reference dataset matching the supplied design assumptions, and
the mean, median, and 95% interval of partial eta squared across successful
simulations. Because power in an unbalanced design depends on the allocation,
means, variances, covariance structure, tested term, and sums-of-squares type,
these effect-size summaries describe this exact design rather than a separate
target effect size.

### Adding factors and levels

You can add factors and levels as needed, and specify any term of interest. For, example if we want to add a between condition with 3 levels, and we are interested in the 3-way interaction, we can do:

```{r complex, eval=FALSE}
power_n(
  between = c(cond = 2, age = 3), # cond has 2 levels, age has 3 levels
  within = c(stim = 4), # stim has 4 levels
  term = "cond:stim:age",
  target_pes = 0.14,
  alpha = 0.05,
  power = 0.90,
  n_sims = 1000, # use 5000+ for a more precise estimate
  seed = 123 # for reproducibility
)
```

## Simulate a power curve

You might want to see how power changes across a range of sample sizes. `power_curve()` simulates power across a range of sample sizes, which you can specify with `n_range`. The result is a tidy data frame that you can plot with `plot_power_curve()`.

```{r curve-fixed-code, eval=FALSE}
pc <- power_curve(
  between = c(cond = 2),
  within = c(stim = 2),
  term = "cond:stim",
  target_pes = 0.14,
  n_range = c(16, 20, 23, 28), # n per between-subject cell
  n_sims = 1000,
  seed = 123
)
pc
```

```{r curve-fixed-output, echo=FALSE}
pc <- vignette_results$curve
pc
```

```{r plot-fixed}
plot_power_curve(
  pc,
  power_lines = c(.80, .90) # adds horizontal lines at 80% and 90% power
)
```

## Nonspherical designs

By default, the balanced simulation functions `power_n()`, `power_curve()`,
`power_achieved()`, and `power_sensitivity()` use a standard deviation of 1 for
every within-subject cell and a correlation of 0.5 between every pair. All four
accept a `within_covariance()` object when repeated measurements have different
standard deviations or correlations. The simulation-free `_calc()` functions
instead accept a planned `epsilon` value.

`power_unbalanced()` is the exception: its standard deviations are supplied in
`cell_design()`, so its `covariance` argument accepts an
`unbalanced_covariance()` object containing correlations only.

The within-subject cell names combine factor levels with underscores. For
`within = c(time = 3, condition = 2)`, the cells are
`time1_condition1`, `time1_condition2`, `time2_condition1`,
`time2_condition2`, `time3_condition1`, and `time3_condition2`. Correlation
names join a pair of cells with `:`.

Define the covariance assumptions separately so they remain easy to read and
check:

```{r covariance-specification, eval=FALSE}
covariance <- within_covariance(
  default_sd = 1,
  default_correlation = 0.5,
  standard_deviations = c(
    "time3_condition2" = 1.2
  ),
  correlations = c(
    "time1_condition1:time1_condition2" = 0.6,
    "time2_condition1:time2_condition2" = 0.7
  )
)
```

Unlisted measurements use `default_sd`, and unlisted pairs use
`default_correlation`. Feed the resulting specification into `power_n()`:

```{r nonspherical-power, eval=FALSE}
power_n(
  within = c(time = 3, condition = 2),
  term = "time:condition",
  target_pes = 0.14,
  power = 0.90,
  n_sims = 1000,
  covariance = covariance,
  seed = 123
)
```

The custom covariance structure is used to calibrate the target effect and to
generate every simulated dataset. For the tested term, the package also
projects the covariance matrix into the term's within-subject contrast space
and calculates a population Greenhouse--Geisser epsilon. `power_calc` applies
that epsilon to its numerator degrees of freedom, denominator degrees of
freedom, and noncentrality parameter. This is equivalent to supplying the
derived value explicitly as `power_n_calc(epsilon = ...)`.

`power_sim` reflects the custom data-generating covariance, but its simulated
ANOVA currently uses the ordinary univariate p-value rather than a
sample-estimated Greenhouse--Geisser or Huynh--Feldt correction.

## Advanced options

### Run simulations in parallel

For larger simulation runs, set `parallel = TRUE`. If you do not set `cores`,
`anovapowersim` uses one fewer than the number of available cores and prints a
message with the chosen count. Set `cores` explicitly when you want a fixed
number of cores.

```{r parallel, eval=FALSE}
power_curve(
  between = c(cond = 2),
  within = c(stim = 2),
  term = "cond:stim",
  target_pes = 0.14,
  n_range = c(16, 20, 23, 28),
  n_sims = 5000,
  parallel = TRUE,
  cores = 4,
  seed = 123
)
```

### Match the G\*Power convention

By default, `anovapowersim` calibrates the simulated cell means so the empirical
reference dataset has the requested partial eta squared under the fitted
`stats::aov()` model. This corresponds to the fitted ANOVA denominator-df
noncentrality convention.

Set `gpower = TRUE` when you want the G\*Power-style convention (when using the 'as in Cohen (1988) option for within-subjects designs)
`lambda = total_n * f^2`. 

```{r gpower-adaptive-code, eval=FALSE}
power_n(
  between = c(cond = 2),
  within = c(stim = 4),
  term = "cond:stim",
  target_pes = 0.14,
  alpha = 0.05,
  power = 0.90,
  n_sims = 1000,
  seed = 123,
  gpower = TRUE
)
```

```{r gpower-adaptive-output, echo=FALSE}
vignette_results$gpower_adaptive
```
