---
title: "Covariance and nonsphericity"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Covariance and nonsphericity}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message=FALSE}
library(anovapowersim)
```

The balanced simulation functions `power_n()`, `power_curve()`,
`power_achieved()`, and `power_sensitivity()` use a common standard deviation
of 1 and a correlation of 0.5 between every within-subject pair by default.
All four accept `within_covariance()` to change that common SD or specify
pair-specific correlations.

## Name the within-subject cells

Cell names combine within-factor levels with underscores. For
`within = c(time = 3, condition = 2)`, the expected names are
`time1_condition1`, `time1_condition2`, `time2_condition1`,
`time2_condition2`, `time3_condition1`, and `time3_condition2`.

Correlation names join a pair of cells with `:`.

```{r covariance-specification, eval=FALSE}
covariance <- within_covariance(
  sd = 1,
  default_correlation = 0.5,
  correlations = c(
    "time1_condition1:time1_condition2" = 0.6,
    "time2_condition1:time2_condition2" = 0.7
  )
)
```

Every measurement uses the common `sd`; unlisted pairs use
`default_correlation`. Pair-specific correlations can produce nonsphericity
even though all diagonal variances are equal.

The package warns whenever these defaults are actually needed:

- Omitting `covariance` from a balanced simulation uses `sd = 1` and a
  correlation of `0.5` for every within-subject pair.
- Calling `within_covariance()` without `sd` uses the common `sd = 1`.
- If `correlations` does not name every pair, the warning reports how many
  pairs are undefined. `default_correlation` fills only those undefined pairs;
  correlations explicitly listed by the user are never replaced.

A raw covariance matrix defines both variances and correlations directly and
does not use these defaults. Its diagonal variances must all be equal.

## Use the covariance specification

Pass the object to any balanced simulation function:

```{r nonspherical-power, eval=FALSE}
power_n(
  within = c(time = 3, condition = 2),
  term = "time:condition",
  target_pes = 0.14,
  power = 0.90,
  n_sims = 5000,
  covariance = covariance,
  seed = 123
)
```

The custom covariance structure calibrates the target effect and generates
every simulated dataset. The package projects the covariance into the tested
term's contrast space and calculates a population Greenhouse--Geisser epsilon.
`power_calc` applies this epsilon to its degrees of freedom and noncentrality.

When the population epsilon is below 1, `power_sim` uses each simulated
dataset's sample-estimated Greenhouse--Geisser-adjusted p-value for sums-of-
squares type II or III. Type I tests remain uncorrected and issue a warning.
Huynh--Feldt corrections are not currently implemented.

## Calculation-only and unbalanced designs

Calculation-only functions accept a planned `epsilon` rather than a
`within_covariance()` object; see [Calculated power](calculated-power.html).

For `power_unbalanced()`, the common SD and correlations belong in
`unbalanced_covariance()`; individual cells specify only sample sizes and
means.
See [Power for unbalanced designs](unbalanced-designs.html).
