---
title: "Power for unbalanced designs"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Power for unbalanced designs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message=FALSE}
library(anovapowersim)
```

`power_unbalanced()` is experimental and available only in the development
version of `anovapowersim`. Use it when the exact allocation, population means,
and standard deviations are already known. It simulates one fixed design; it
does not search over sample sizes or return calculated power.

## Define the cells

Define every cell with `cell_design()`. Factor values identify the cell, `n`
is the number of subjects in its between-subject group, `m` is its population
mean, and `sd` is its marginal population standard deviation. Name repeated-
measures factors in `within`; all remaining factors are between-subject factors.

For repeated-measures designs, the same `n` must appear on every within-subject
row belonging to a given between-subject group.

```{r unbalanced-design, eval=FALSE}
unbalanced_design <- cell_design(
  group = "control",   time = "pre",  n = 22, m = 10.0, sd = 2.0,
  group = "control",   time = "post", n = 22, m = 11.0, sd = 2.2,
  group = "treatment", time = "pre",  n = 31, m = 10.1, sd = 2.4,
  group = "treatment", time = "post", n = 31, m = 12.4, sd = 2.8,
  within = "time"
)
```

The means define the complete effect pattern, including main effects and
interactions. Standard deviations can differ across groups and within-subject
cells.

Every factor must have at least two observed levels, and every factor-level
combination must be defined exactly once. If cells are missing,
`cell_design()` reports which ones. Supply `default_n`, `default_m`, and
`default_sd` together to fill missing cells automatically:

```{r unbalanced-default-fill, eval=FALSE}
cell_design(
  group = "control",   time = "pre",  n = 22, m = 10.0, sd = 2.0,
  group = "control",   time = "post", n = 22, m = 11.0, sd = 2.2,
  group = "treatment", time = "pre",  n = 31, m = 10.1, sd = 2.4,
  within = "time",
  default_n = 31, default_m = 12.4, default_sd = 2.8
)
```

## Define within-subject correlations

For repeated measures, use `unbalanced_covariance()` to define correlations.
It deliberately has no standard-deviation argument: `power_unbalanced()`
combines its correlations with the cell-specific `sd` values to construct a
covariance matrix for each between-subject group.

With one within-subject factor, pair names use its levels, such as
`"pre:post"`.

```{r unbalanced-power, eval=FALSE}
power_unbalanced(
  design = unbalanced_design,
  term = "group:time",
  covariance = unbalanced_covariance(
    default_correlation = 0.5,
    correlations = c("pre:post" = 0.7)
  ),
  n_sims = 5000,
  parallel = TRUE,
  seed = 123
)
```

## Interpret the result

The result reports simulated power, partial eta squared from a deterministic
reference dataset matching the design assumptions, and the mean, median, and
95% interval of partial eta squared across successful simulations. These
effect-size summaries describe the exact allocation, means, variances,
correlations, tested term, and sums-of-squares type supplied by the user.

## Multiple within-subject factors

Name every repeated-measures factor in `cell_design()` and join their level
combinations with underscores in correlation-pair names:

```{r unbalanced-multi-within, eval=FALSE}
multi_within_design <- cell_design(
  group = "A", time = "pre",  cond = "control", n = 10, m = 0.0, sd = 1,
  group = "A", time = "pre",  cond = "treat",   n = 10, m = 0.5, sd = 1,
  group = "A", time = "post", cond = "control", n = 10, m = 0.2, sd = 1,
  group = "A", time = "post", cond = "treat",   n = 10, m = 1.0, sd = 1,
  group = "B", time = "pre",  cond = "control", n = 15, m = 0.0, sd = 1,
  group = "B", time = "pre",  cond = "treat",   n = 15, m = 0.6, sd = 1,
  group = "B", time = "post", cond = "control", n = 15, m = 0.3, sd = 1,
  group = "B", time = "post", cond = "treat",   n = 15, m = 1.4, sd = 1,
  within = c("time", "cond")
)

power_unbalanced(
  design = multi_within_design,
  term = "group:time:cond",
  covariance = unbalanced_covariance(
    correlations = c("pre_control:post_control" = 0.6)
  ),
  n_sims = 5000,
  seed = 123
)
```
